Practical Optional Types for Clojure

### 2015-12-30 formal-model.tex
more TODO's

Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
in T-App the operand \ep{} = \makelocal{d} is checked with
T-Local as
\begin{equation}
\judgementselfrewrite{\propenv{}}
                     {\makelocal{d}}
                     {\Top}
                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
                                {\isprop{\falsy}{\makelocal{d}}}}
                     {\makelocal{d}}
\end{equation}
which encodes the type, proposition, and object information
about variables.

Finally, the T-App rule substitutes the operand's object \objectp{}
for the parameter \x{} in the latent type, propositions, and object.
\begin{equation}
\judgementselfrewrite{\propenv{}}
  {\appexp{\numberhuh{}}{\makelocal{d}}}
  {\Boolean{}}
  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
             {\notprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
\end{equation}
To demonstrate, the `then' proposition---in T-App {\replacefor {\thenprop{\prop{}}} {\objectp{}} {\x{}}}---substitutes
the latent `then' proposition of \constanttype{\numberhuh{}} with 
\makelocal{d}, giving
{\replacefor {\isprop{\Number{}}{\x{}}} {\makelocal{d}} {\x{}}} =
{\isprop{\Number{}}{\makelocal{d}}}.

@@

Next, show how to check
$
{\extendmultiexp {hi_0} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
$.
%
The expression 
{\extendmultiexp {\e{m}} {\e{v}} {\e{f}}} creates a new multimethod
that extends multimethod \e{m}'s dispatch table, mapping dispatch value
\e{v} to method \e{f}. The T-DefMulti typing rule
checks \e{m} is a multimethod with dispatch function type \t{d},
then calculates the extra information we know based on the current
dispatch value {\thenprop{\proppp{}}}, which is assumed when checking the method
body.
%$$
%    \TDefMethod{}
%$$
Our example checks with \e{m} being of type
$
{\MultiFntype {\ArrowTwo{\x{}}{\Keyword}{\String}}
              {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}}
$
with \objectp{} = {\x{}} and \t{v} = \Value{\makekw{en}}. Then {\thenprop{\proppp{}}} = 
{\isprop {\Value{\makekw{en}}}{\x{}}} by
$
\isacompare{}{\x{}}{\Value{\makekw{en}}}
{\filterset {\isprop {\Value{\makekw{en}}}{\x{}}}{\notprop{\Value{\makekw{en}}}{\x{}}}}
$.
Since \t{} = \Keyword{}, we check the method body with
$
\judgement{{\isprop{\Keyword}{\x{}}},{\isprop {\Value{\makekw{en}}}{\x{}}}}
  {\makestr{hello}}
  {\String}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$.
Finally from the interface type \t{m}, we know \thenprop{\prop{}} = \elseprop{\prop{}} = \topprop{},
and \object{} = \emptyobject{}, which also agrees with the method body, above.

>>>>

Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
in T-App the operand \ep{} = \makelocal{d} is checked with
T-Local as
\begin{equation}
\judgementselfrewrite{\propenv{}}
                     {\makelocal{d}}
                     {\Top}
                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
                                {\isprop{\falsy}{\makelocal{d}}}}
                     {\makelocal{d}}
\end{equation}
which encodes the type, proposition, and object information
about variables. The proposition {\notprop{\falsy}{\makelocal{d}}}
says ``it is not the case that variable {\makelocal{d}} is of type {\falsy}'';
{\isprop{\falsy}{\makelocal{d}}} says ``{\makelocal{d}} is of type {\falsy}''.

Finally, the T-App rule substitutes the operand's object \objectp{}
for the parameter \x{} in the latent type, propositions, and object. The proposition
{\isprop{\Number{}}{\makelocal{d}}} says ``{\makelocal{d}} is of type {\Number{}}'';
{\notprop{\Number{}}{\makelocal{d}}} says ``it is not the case that {\makelocal{d}}
is of type {\Number{}}''. The object {\makelocal{d}} is the symbolic representation
of what the expression {\makelocal{d}} evaluates to.
\begin{equation}
\judgementselfrewrite{\propenv{}}
  {\appexp{\numberhuh{}}{\makelocal{d}}}
  {\Boolean{}}
  {\filterset{\isprop{\Number{}}{\makelocal{d}}}
             {\notprop{\Number{}}{\makelocal{d}}}}
  {\emptyobject{}}
\end{equation}
To demonstrate, the `then' proposition---in T-App {\replacefor {\thenprop{\prop{}}} {\objectp{}} {\x{}}}---substitutes
the latent `then' proposition of \constanttype{\numberhuh{}} with 
\makelocal{d}, giving
{\replacefor {\isprop{\Number{}}{\x{}}} {\makelocal{d}} {\x{}}} =
{\isprop{\Number{}}{\makelocal{d}}}.

@@

Next, we show how to check
$
{\extendmultiexp {hi_0} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
$.
%
The expression 
{\extendmultiexp {\e{m}} {\e{v}} {\e{f}}} creates a new multimethod
that extends multimethod \e{m}'s dispatch table, mapping dispatch value
\e{v} to method \e{f}. The T-DefMulti typing rule
checks \e{m} is a multimethod with dispatch function type \t{d},
then calculates the extra information we know based on the current
dispatch value {\thenprop{\proppp{}}}, which is assumed when checking the method
body.
%$$
%    \TDefMethod{}
%$$
Our example checks with \e{m} being of type
$
{\MultiFntype {\ArrowTwo{\x{}}{\Keyword}{\String}}
              {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}}
$
with \objectp{} = {\x{}} and \t{v} = \Value{\makekw{en}}. Then {\thenprop{\proppp{}}} = 
{\isprop {\Value{\makekw{en}}}{\x{}}} by
$
\isacompare{}{\x{}}{\Value{\makekw{en}}}
{\filterset {\isprop {\Value{\makekw{en}}}{\x{}}}{\notprop{\Value{\makekw{en}}}{\x{}}}}
$.
Since \t{} = \Keyword{}, we check the method body with
$
\judgement{{\isprop{\Keyword}{\x{}}},{\isprop {\Value{\makekw{en}}}{\x{}}}}
  {\makestr{hello}}
  {\String}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$.
Finally from the interface type \t{m}, we know \thenprop{\prop{}} = \elseprop{\prop{}} = \topprop{},
and \object{} = \emptyobject{}, which also agrees with the method body, above.
Notice the overall type of a \defmethodliteral{} is the same as its first subexpression \e{m}.

### 2015-12-30 overview.tex
fix italic gamma
more TODO's

(ann enc-keypair [RawKeyPair -> EncKeyPair])
(defn enc-keypair [kp]
  (assoc (dissoc kp :priv) :enc-priv (encrypt (:priv kp))))
\end{minted}
\label{example:circleci}
\end{exmp}

%(ann enc-keypair [RawKeyPair -> EncKeyPair])
%(defn enc-keypair [{pk :priv :as kp}] ; original map is kp
%  (assoc (dissoc kp :priv)       ; remove unencrypted private key
%         :enc-priv (encrypt pk))) ; add encrypted private key

%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
As \clj{EncKeyPair} is fully specified, we must remove extra keys like \clj{:priv}.
%
\begin{minted}{clojure}
(defn enc-keypair-bad [kp] ; Type error: :priv disallowed
  (assoc kp :enc-priv (encrypt (:priv kp))))
\end{minted}

>>>>

(ann enc-keypair [RawKeyPair -> EncKeyPair])
(defn enc-keypair [kp]
  (assoc (dissoc kp :priv) :enc-priv (encrypt (:priv kp))))
\end{minted}
\label{example:circleci}
\end{exmp}
%
%(ann enc-keypair [RawKeyPair -> EncKeyPair])
%(defn enc-keypair [{pk :priv :as kp}] ; original map is kp
%  (assoc (dissoc kp :priv)       ; remove unencrypted private key
%         :enc-priv (encrypt pk))) ; add encrypted private key
%
%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
As \clj{EncKeyPair} is fully specified, we remove extra keys like \clj{:priv}
via \clj{dissoc}, which returns a new map that is the first argument without the
entry named by the second argument. Notice removing \clj{dissoc} causes a type error.
%
\begin{minted}{clojure}
(defn enc-keypair-bad [kp] ; Type error: :priv disallowed
  (assoc kp :enc-priv (encrypt (:priv kp))))
\end{minted}
